A Data-Driven Framework for Neural Field Modeling

### 2011-01-07 ltx/first_order_framework/response_to_reviews.tex

        We first address the reviewer's main concerns, followed by the reviewer's additional comments.

    \subsection{Main Concerns}
    It would be useful to mention the above (now below) explicitly in the introduction and try to address them in later sections of the paper.

    \begin{enumerate}
        \item Do not use real data for the validation of their approach

	\emph{The authors appreciate that the only way to truly validate the proposed framework is with real data. However, before validating the method on real data we believe that it is essential to first validate the estimation method using accepted models and synthetic data, where the true and estimated parameters can be compared. In an effort to make the method employed in the paper more explicit, we have added ``To illustrate the estimation framework, data is generated using the neural field equations incorporating modeled sensors enabling a comparison between the estimated and true parameters'' to the introduction. Another paper demonstrating the application of the method with data recorded from a Utah array implanted in an epilepsy surgical patient is currently in preparation. We have also added that validating the framework on real data is noted as future work in the conclusion.}

@@

        \item Assume that a few kernel basis functions accurately represent the actual connectivity kernel, which might not be the case when using real data;

	\emph{The examples presented in the paper use the Mexican hat connectivity kernel. The framework is also valid for other connectivity kernels provided they are spatially homogeneous and that they can be approximated by a set of weighted isotropic basis functions. The mathematical formulation does not place any further restrictions on the form of the spatial mixing kernel. We have added statements to the methods section (after equation~13) to be more explicit in describing subtleties of the method.}

@@

        \item The use of Euler discretisation and how this affects the system's behavior.

		\emph{This point was made by a reviewer from our PLoS attempt. The reviewer stated: ``1. The authors might want to reference prior work in using Kalman
		Filtering for the estimation of neural mass models from data [1-3],
		[13]. In fact from those papers stressed the importance of carrying out
		a high order discretization of the differential equations, which should
		be in fact treated as stochastic. In equation (S1.1) of the extended
		derivation, it is shown that an Euler approximation is taken which may
		even change the dynamical properties of the system. The authors should
		discuss this issue.'' ALL THESE PAPERS USE a method called local linearisation approach for discretisation.}

		\emph{We should add some more to the modeling assumptions section to address this one. I have begun to talk about the LL method.}

        \item The derivation of Equations (30) and (31) and make them more transparent for the general audience.

\emph{To address this point, we have added an extra appendix with a derivation of the sampling theorem with the appropriate cross-referencing to the main text. An effort was made to keep this derivation minimal and as clear as possible. We have also added the references to relevant part of the text (that should have been there before, as the reviewer has correctly pointed out). We felt that it was worth adding the extra appendix to provide a clear derivation that is contextually and notationally relevant to this paper.}

@@

    \section{Reviewer 2}

    We thank the reviewer for their encouraging review. Each point raised is dealt with in turn.

\begin{enumerate}
    \item The handling of the (to borrow a term from numerical weather forecasting) `sub-grid scale physics' is inadequate for nervous systems (see, e.g., equation 8 [The nonlinear IntegroDifference Equation]). I do very much appreciate the spectral estimation of spatial and temporal scales. But the model errors are lumped into an iid model error term that is not realistic. For instance, the flow of potassium in the extracellular spaces has a known dynamic that will affect network excitability, generating important sub-grid dynamics that is anything but iid. See, for instance, the example in Ullah et al PLoS Comp Biol 2010 where without such dynamics, state tracking and parameter estimation for a neural system fails. A beautiful overview of sub-grid effects in data assimilation is found in Kalnay's 2003 book. I would very much value the authors comments on these issues. I guess were I to respond to this question, that I might consider extending the model error term to an autoregressive model. But I will let the authors answer this question.\\

	\emph{To address this point, we have added an extra section to the discussion, `Modeling Assumptions'. Within this section we discuss model mismatch and what we think is a good way to deal with it.}
	\emph{\mike{We would like to directly address the idea of extending the error term to a higher order model. The reason we have not performed this kind of extension in the paper is that we have attempted to remain as close as possible to the original Wilson and Cowan model, to show that such models can be inferred from data. While we are considering various model extensions, that of higher order temporal dynamics are particularly intriguing due to the additional richness of signal we can represent. Done properly though, this will introduce additional connectivity kernels (one for each temporal lag in the AR process), and would hence imply something about the connectivity in the field that we would need to relate to an extension in the underlying field model or its discretisation.}}

@@

\item What is the relation of the substitution of the temporal and spatial basis functions as employed in this paper to Galerkin's projection? How do the two techniques differ? If this is close, the non-local case for a Galerkin projection (typically used in diffusive processes) has not, to my knowledge, been done. Certainly not in a way that is obvious for neural field theory. Perhaps this is my ignorance. This seems important to clarify for the rest of us.

\emph{The relationship to our basis function decomposition to Galerkin's projection has been added to the derivation of the state-space model. }

\item In the isotropic and translationally invariant case examined here, would a Fourier basis for temporal and spatial structure have been a formally correct alternative?\\

\emph{Fourier basis functions would be a correct alternative to the case examined in this paper. In fact Fourier basis functions or any other orthonormal basis functions would simplify the state-space representation of the model, since $\Gamma$ (defined in equation (20)) of the manuscript would be the identity matrix. However, we have chosen Gaussian basis functions due to their semi-compact support.} %By choosing Gaussian basis functions the state-space representation of the model remains in its generic form which is also correct in the case of orthonormal basis functions.}

@@

\item In equation 29, what would a reasonable $\boldsymbol{\nu}_c$ be for a real neuronal system (mammalian) given the number of neurons and scale of electrodes? I am concerned that it might be much higher than any of us wish to consider. One would perhaps be saved by the filter characteristics of brain (e.g. Phys. Rev. E 73, 051911 (2006)).

\emph{The mean spatial frequency content is seems to be actually lower than one might first think. We have added `From the work of (Freeman et al, 2000), the expected bandwidth of the human neocortex is approximately 0.1-0.4 cycles/mm which requires a sensor spacing of approximately 1.25 mm to avoid distortions due to aliasing' to the Frequency Analysis section of the paper to clear this up. We have also added to the discussion. Note, we expect the spatial bandwidth of the cortex to be governed by the connectivity kernel, since it is equivalent to a spatial low-pass filter acting on the firing rates.}

@@

\emph{This point is clarified by adding the connectivity kernel components superimposed by the resulting Mexican Hat basis in Figure 1. of the manuscript.}

 \begin{figure*}[!ht]
 \begin{center}
 \includegraphics{./Graph/pdf/fig1.pdf} 
 \end{center}
 \caption{}
 \label{fig:Figure1}
 \end{figure*}

\item A few comments on how to adapt the smoother for real time implementation would be helpful. I presume that one would perform the smoothing during a training phase. But one would also wish to adapt in real time as observation or control scenarios progressed. A few words of perspective would be invaluable to help others translate the author's wisdom on this method to other applications.

\emph{A few comments have been added to the `Extensions to the Framework' section of the discussion.}

@@

\item Figure 4. Should the axes be indicated as spatial frequency rather than just Hz?

\emph{Axes of Figure 4. (and also Figure 5.) are now changed to spatial frequency. The unit of the spatial frequency within the text is changed to cycles/mm.}

\item Page 45, Appendix D, Par 1, line 6: should `transform of a n-' should be `transform of an n-' ? I am grammatically challenged perhaps.

\emph{We stand corrected. We have changed the text to `an n-dimensional'.}

### 2011-01-07 ltx/first_order_framework/LaTeX_Manuscript.tex
edits to discussion

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace} 
\usepackage{color}

% \newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}}
\newcommand{\dean}[1]{\textcolor{red}{#1}}
\newcommand{\parham}[1]{\textcolor{blue}{#1}}

@@

Each connectivity basis function of the Mexican hat can, individually, be considered a layer in the Wilson and Cowan model. Rearranging the order of the spatial convolution in equation~\ref{DiscreteTimeModel} and substituting in equations~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} \dean{to the right hand side} we get the approximation of the dynamic field
% 	\label{reduced continuous model}
% 	\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_{t+1} &= T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
% 	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}). 
	\label{reduced continuous model}
	v_{t+1}\left(\mathbf{r}\right) &\approx T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}). 
Next we cross-multiply equation~\ref{reduced continuous model} by $\boldsymbol{\phi}(\mathbf{r})$ and integrate over the spatial domain, $\Omega$, to get 
	\label{eq:decompRHS}
 	\int_\Omega \boldsymbol{\phi}&\left(\mathbf{r}\right) v_{t+1}\left(\mathbf{r}\right)d\mathbf{r}  \nonumber \\
	&\approx T_s \int_\Omega \boldsymbol{\phi} (\mathbf{r}) \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}') \mathbf{x}_t ) \, d\mathbf{r}'d\mathbf{r}\boldsymbol{\theta} \nonumber \\ 
	&+ \xi\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r} } \mathbf{x}_t + \int_\Omega{\boldsymbol{\phi} (\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}. 
\dean{The solution of this equation for $\mathbf{x}_t$ is equivalent to finding a Galerkin projection, where the field basis functions would be considered the Galerkin projection functions and the states would be the amplitude coefficients. In the next step to forming the state-space model we substitute the field decomposition into the left hand side of equation~\ref{eq:decompRHS} giving}
	\label{eq:StartofReduction}
 	\int_\Omega \boldsymbol{\phi}&\left(\mathbf{r}\right) \boldsymbol{\phi}^{\top}\left(\mathbf{r}\right)\mathbf{x}_{t+1}d\mathbf{r}  \nonumber \\
	&= T_s \int_\Omega \boldsymbol{\phi} (\mathbf{r}) \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}') \mathbf{x}_t ) \, d\mathbf{r}'d\mathbf{r}\boldsymbol{\theta} \nonumber \\ 
	&+ \xi\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r} } \mathbf{x}_t + \int_\Omega{\boldsymbol{\phi} (\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}. 
Now defining the matrix
\begin{equation}\label{eq:DefGamma}
	\boldsymbol{\Gamma} \triangleq \int_\Omega {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi} ^{\top}\left(\mathbf{r}\right) \, d\mathbf{r}}, 
and substituting it into equation~\ref{eq:StartofReduction} and cross-multiplying by $\boldsymbol{\Gamma}^{-1}$ gives
    \label{eq:ReducedForm}
	 \mathbf{x}_{t+1} &= T_s\boldsymbol{\Gamma}^{-1}
	 \int_\Omega \boldsymbol{\phi}(\mathbf{r}) 
	 \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}')f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t)d\mathbf{r}' d\mathbf{r} \boldsymbol{\theta}  \nonumber \\
	& + \xi\mathbf{x}_t + \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}.
The analytic derivation of the inner product of \emph{n}-dimensional Gaussians, which was used for calculating $\mathbf\Gamma$, is provided in \ref{App:InnerProdOfGaussians}. \dean{Alternative field basis functions may be used to form the decomposition and if the basis is orthonormal, such as a Fourier basis, then $\Gamma$ is simply the identity matrix.} Equation~\ref{eq:ReducedForm} can be simplified by exploiting the isotropy of the connectivity kernel basis functions where
	\psi_i (\mathbf{r}-\mathbf{r}') = \psi_i (2\mathbf{c}_i+\mathbf{r}'-\mathbf{r})
and $\mathbf{c}_i$ is the center of the $i^{th}$ basis function of the kernel decomposition. To make the simplification we define
\begin{equation}\label{eq:DefPsi}
	\left[ \boldsymbol\Psi(\mathbf{r}')\right]_{:i}  \triangleq T_s\boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\psi_i (2\mathbf{c}_i+\mathbf{r}'-\mathbf{r})\textrm{d}\mathbf{r}},
where $\left[ \boldsymbol\Psi(\mathbf{r}')\right]_{:i}$ denotes the $i^{th}$ column of $\boldsymbol{\Psi}(\mathbf{r}')$ which is a $n_x \times n_{\theta}$ matrix, where $n_x$ is the number of basis functions (states) and $n_{\theta}$ is the number of connectivity kernel basis functions. This matrix is pre-defined analytically (see \ref{ap:ConvOfGaussians} for analytic convolution of two Gaussians), since it is constant with respect to the states and the unknown parameters. Substituting $\boldsymbol{\Psi}(\mathbf{r}')$ into equation~\ref{eq:ReducedForm} gives
	\mathbf{x}_{t+1} &= \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) \, d\mathbf{r}' \boldsymbol{\theta} \nonumber \\ &+ \xi\mathbf{x}_t 
+ \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r}) \, d\mathbf{r}}.
This substitution swaps the convolution in equation~\ref{reduced continuous model}, which is computationally demanding, with the convolution in equation~\ref{eq:DefPsi}, which is pre-defined analytically. This is of great importance as it lowers the computational demands of the estimation algorithm and provides a dramatic increase in estimation speed.

@@

The additive form of the unscented Rauch-Tung-Striebel smoother (URTSS) \citep{Sarkka2010} was used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF) \citep{Julier1997, Merwe2003} in a forward step to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward step to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{align}\label{eq:sigmapoints1}
	\mathcal X_{0}&=\mathbf{\bar x} \\
	\mathcal X_{i}&= \mathbf{\bar x}+\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i, \quad i=1, \dots, n_x \\
	\mathcal X_{i}&=\mathbf{\bar x}-\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_{i- n_x}, \quad i= n_x+1, \dots, 2n_x 
\end{align}
where $\mathbf{\bar x}$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix for filtering or smoothing, $\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i$ is the $i^{th}$ column of the scaled matrix square root of $\mathbf P_x$, and $n_x$ is the dimension of the state space. The total number of sigma points is $2n_x+1$. The scaling parameter, $\lambda$, is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( n_x+\kappa) - n_x, 
\end{equation}
\parham{(reviewer 2, comment 7:) The positive constant $\alpha$, that determines the spread of the sigma points around $\mathbf{\bar x}$, is typically set arbitrary small (e.g., $10^{-3}$) to minimise higher order effect \citep{Haykin2001}, and  $\kappa$ is an additional scaling parameter which was set to $3-n_x$ in accordance to \citet{Julier2002a}.}

@@

\subsection{Modeling Assumptions}
\dean{(comment 1 and 9 from reviewer 2:) It is an implicit assumption that the neural field model (in equation~\ref{DiscreteTimeModel}) provides an apt description of the cortical dynamics recorded from the subject. Although models of this form are capable of describing a variety of cortical dynamics, there is without doubt, a mismatch between the cortex and the model. Nevertheless, there is a sufficient volume of interesting results from theoretical studies that warrants the data assimilation framework. }

\dean{An advantage in using a lumped-parameter field model over a more detailed mathematical description is that the myriad of parameters that might influence excitability, such as specific ion concentrations, are considered to be lumped into the connectivity kernel coefficients that effectively describe the net system gain. In this way a minimal number of assumptions are made. A major challenge in data assimilation for the mass action of the brain is to make models sufficiently detailed such that the parameters are meaningful, but simple enough to yield clear insights that can be related to theoretical studies (from both neuroscience and engineering perspectives).}

\dean{The framework assumes that the parameters of the model (and brain) are stationary over the estimation period. This is a reasonable assumption if applying the framework to an evoked-potential experiment over a short time period in a controlled environment. However, it is well known that factors that influence excitability are non-stationary over longer time periods (see e.g. \citet{Ullah2010}). Therefore, the estimation framework must be altered if it is to be applied to data over extended periods of time. A possible way to achieve this is to use a more sophisticated model that accounts for longer term time dependancies (higher temporal order). Alternatively, the parameters of the model could be considered time-varying, where changes in excitability might result in different connectivity kernel parameters. A possible way to achieve this would be to employ window-based estimation or fixed-lag, where the estimation is performed on successive windows of data (possibly overlapping windows). The window length will need to be sufficiently short such that the parameters are approximately stationary of the estimation duration, and sufficiently long to achieve a good estimation result. If this trade-off can not be successfully achieved then the level of detail in the model must be increased. A real-time implementation can also be achieved using the UKF \parham{and an online recursive LS estimator (see \citet{Ljung1999} for example)}. }

@@

\section{Conclusion}
This paper has presented a framework for creating data-driven neural field models by developing a method for estimating intracortical connectivity and synaptic dynamics from electrophysiological data. In doing so, we have highlighted important aspects of the experiment design process, demonstrating the relationship between sensor bandwidth and the complexity of the field that can be inferred. We have derived a finite-order state-space representation of the continuum neural field equations that allows inference to be performed efficiently. Importantly, the mathematical formulation that we have presented scales well with the number of sensors. The major contribution of this framework is to formally bridge a gap between experimental and theoretical neuroscience in order to gain a better understanding of the mesoscopic dynamics of the brain. \dean{The validation of the estimation framework on real data is now required, and is the subject of planned future work.}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace} 
\usepackage[usenames,dvipsnames]{color}

% \newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}}
\newcommand{\dean}[1]{\textcolor{red}{#1}}
\newcommand{\parham}[1]{\textcolor{blue}{#1}}
\newcommand{\mike}[1]{\textcolor{Plum}{#1}}

@@

Each connectivity basis function of the Mexican hat can, individually, be considered a layer in the Wilson and Cowan model. Rearranging the order of the spatial convolution in equation~\ref{DiscreteTimeModel} and substituting in equations~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} \dean{to the right hand side} we get the approximation of the dynamic field
% \begin{align}
% 	\label{reduced continuous model}
% 	\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_{t+1} &= T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
% 	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}). 
% \end{align}
\begin{align}
	\label{reduced continuous model}
	v_{t+1}\left(\mathbf{r}\right) &\approx T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}). 
\end{align}
Next we cross-multiply equation~\ref{reduced continuous model} by $\boldsymbol{\phi}(\mathbf{r})$ and integrate over the spatial domain, $\Omega$, to get 
\begin{align}
	\label{eq:decompRHS}
 	\int_\Omega \boldsymbol{\phi}&\left(\mathbf{r}\right) v_{t+1}\left(\mathbf{r}\right)d\mathbf{r}  \nonumber \\
	&\approx T_s \int_\Omega \boldsymbol{\phi} (\mathbf{r}) \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}') \mathbf{x}_t ) \, d\mathbf{r}'d\mathbf{r}\boldsymbol{\theta} \nonumber \\ 
	&+ \xi\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r} } \mathbf{x}_t + \int_\Omega{\boldsymbol{\phi} (\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}. 
\end{align}
\dean{The solution of this equation for $\mathbf{x}_t$ is equivalent to finding a Galerkin projection, where the field basis functions would be considered the projection functions and the states would be the amplitude coefficients. See~\ref{App:Galerkin} for more information.}

In the next step of forming the state-space model we substitute the field decomposition into the left hand side of equation~\ref{eq:decompRHS} giving
\begin{align}
	\label{eq:StartofReduction}
 	\int_\Omega \boldsymbol{\phi}&\left(\mathbf{r}\right) \boldsymbol{\phi}^{\top}\left(\mathbf{r}\right)\mathbf{x}_{t+1}d\mathbf{r}  \nonumber \\
	&= T_s \int_\Omega \boldsymbol{\phi} (\mathbf{r}) \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}') \mathbf{x}_t ) \, d\mathbf{r}'d\mathbf{r}\boldsymbol{\theta} \nonumber \\ 
	&+ \xi\int_\Omega {\boldsymbol{\phi}(\mathbf{r}) \boldsymbol{\phi}^{\top}(\mathbf{r})d\mathbf{r} } \mathbf{x}_t + \int_\Omega{\boldsymbol{\phi} (\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}. 
\end{align}
Now defining the matrix
\begin{equation}\label{eq:DefGamma}
	\boldsymbol{\Gamma} \triangleq \int_\Omega {\boldsymbol{\phi} \left(\mathbf{r}\right)\boldsymbol{\phi} ^{\top}\left(\mathbf{r}\right) \, d\mathbf{r}}, 
\end{equation}
and substituting it into equation~\ref{eq:StartofReduction} and cross-multiplying by $\boldsymbol{\Gamma}^{-1}$ gives
\begin{align}
    \label{eq:ReducedForm}
	 \mathbf{x}_{t+1} &= T_s\boldsymbol{\Gamma}^{-1}
	 \int_\Omega \boldsymbol{\phi}(\mathbf{r}) 
	 \int_\Omega \boldsymbol{\psi}^{\top} (\mathbf{r}-\mathbf{r}')f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t)d\mathbf{r}' d\mathbf{r} \boldsymbol{\theta}  \nonumber \\
	& + \xi\mathbf{x}_t + \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r}) e_t(\mathbf{r}) \, d\mathbf{r}}.
\end{align}
The analytic derivation of the inner product of \emph{n}-dimensional Gaussians, which was used for calculating $\mathbf\Gamma$, is provided in \ref{App:InnerProdOfGaussians}. \dean{Alternative field basis functions may be used to form the decomposition and if the basis is orthonormal, such as a Fourier basis, then $\Gamma$ is simply the identity matrix.} Equation~\ref{eq:ReducedForm} can be simplified by exploiting the isotropy of the connectivity kernel basis functions where
\begin{equation}
	\psi_i (\mathbf{r}-\mathbf{r}') = \psi_i (2\mathbf{c}_i+\mathbf{r}'-\mathbf{r})
\end{equation}
and $\mathbf{c}_i$ is the center of the $i^{th}$ basis function of the kernel decomposition. To make the simplification we define
\begin{equation}\label{eq:DefPsi}
	\left[ \boldsymbol\Psi(\mathbf{r}')\right]_{:i}  \triangleq T_s\boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\psi_i (2\mathbf{c}_i+\mathbf{r}'-\mathbf{r})\textrm{d}\mathbf{r}},
\end{equation}
where $\left[ \boldsymbol\Psi(\mathbf{r}')\right]_{:i}$ denotes the $i^{th}$ column of $\boldsymbol{\Psi}(\mathbf{r}')$ which is a $n_x \times n_{\theta}$ matrix, where $n_x$ is the number of basis functions (states) and $n_{\theta}$ is the number of connectivity kernel basis functions. This matrix is pre-defined analytically (see \ref{ap:ConvOfGaussians} for analytic convolution of two Gaussians), since it is constant with respect to the states and the unknown parameters. Substituting $\boldsymbol{\Psi}(\mathbf{r}')$ into equation~\ref{eq:ReducedForm} gives
\begin{align}
	\mathbf{x}_{t+1} &= \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) \, d\mathbf{r}' \boldsymbol{\theta} \nonumber \\ &+ \xi\mathbf{x}_t 
+ \boldsymbol{\Gamma}^{-1} \int_\Omega{\boldsymbol{\phi}(\mathbf{r})e_t(\mathbf{r}) \, d\mathbf{r}}.
\end{align}
This substitution swaps the convolution in equation~\ref{reduced continuous model}, which is computationally demanding, with the convolution in equation~\ref{eq:DefPsi}, which is pre-defined analytically. This is of great importance as it lowers the computational demands of the estimation algorithm and provides a dramatic increase in estimation speed.

@@

The additive form of the unscented Rauch-Tung-Striebel smoother (URTSS) \citep{Sarkka2010} was used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF) \citep{Julier1997, Merwe2003} in a forward step to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward step to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. The sigma points, $\mathcal X_i$, are calculated using the unscented transform as follows:
\begin{align}\label{eq:sigmapoints1}
	\mathcal X_{0}&=\mathbf{\bar x} \\
	\mathcal X_{i}&= \mathbf{\bar x}+\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i, \quad i=1, \dots, n_x \\
	\mathcal X_{i}&=\mathbf{\bar x}-\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_{i- n_x}, \quad i= n_x+1, \dots, 2n_x 
\end{align}
where $\mathbf{\bar x}$ represents either $\hat{\mathbf x}_t^{f}$ or $\hat{\mathbf x}_t^{b}$, $\mathbf{P}_x$ is the corresponding covariance matrix for filtering or smoothing, $\left(\sqrt{( n_x + \lambda)\mathbf P_x}\right)_i$ is the $i^{th}$ column of the scaled matrix square root of $\mathbf P_x$, and $n_x$ is the dimension of the state space. The total number of sigma points is $2n_x+1$. The scaling parameter, $\lambda$, is defined as 
\begin{equation}\label{eq:sigmapoints3}
	\lambda=\alpha^2( n_x+\kappa) - n_x, 
\end{equation}
\parham{(reviewer 2, comment 7:) The positive constant $\alpha$ that determines the spread of the sigma points around $\mathbf{\bar x}$ is typically set arbitrary small (e.g., $10^{-3}$) to minimise higher-order effects \citep{Haykin2001}, and  $\kappa$ is an additional scaling parameter which was set to $3-n_x$ in accordance to \citet{Julier2002a}.}

@@

\subsection{Modeling Assumptions}
\dean{(comment 1 and 9 from reviewer 2:) It is an implicit assumption that the neural field model (in equation~\ref{DiscreteTimeModel}) provides an apt description of the cortical dynamics recorded from the subject. Although models of this form are capable of describing a variety of cortical dynamics, there is without doubt, a mismatch between the cortex and the model. Nevertheless, there is a sufficient volume of interesting results from theoretical studies using the WC field equations that warrants the data assimilation framework. }

\dean{An advantage in using a lumped-parameter field model over a more detailed mathematical description is that the myriad of parameters that might influence excitability, such as specific ion concentrations, are considered to be lumped into the connectivity kernel coefficients that effectively describe the net system gain. In this way a minimal number of assumptions are made. A major challenge in data assimilation for the mass action of the brain is to make models sufficiently detailed, such that the parameters are meaningful, but simple enough to yield clear insights that can be related to theoretical studies (from both neuroscience and engineering perspectives).}

\dean{The framework assumes that the parameters of the model (and brain) are stationary over the estimation period. This is a reasonable assumption if applying the framework to an evoked-potential experiment over a short time period in a controlled environment. However, it is well known that factors that influence excitability are non-stationary over longer time periods (see e.g. \citet{Ullah2010}). Therefore, the estimation framework must be altered if it is to be applied to data over extended periods of time.}

% A possible way to achieve this is to use a more sophisticated model that accounts for longer term time dependancies (higher temporal order). Alternatively, the parameters of the model could be considered time-varying, where changes in excitability might result in different connectivity kernel parameters. A possible way to achieve this would be to employ window-based estimation or fixed-lag, where the estimation is performed on successive windows of data (possibly overlapping windows). The window length will need to be sufficiently short such that the parameters are approximately stationary of the estimation duration, and sufficiently long to achieve a good estimation result. If this trade-off can not be successfully achieved then the level of detail in the model must be increased. A real-time implementation can also be achieved using the UKF \parham{and an online recursive LS estimator (see \citet{Ljung1999} for example)}. }

@@

\mike{The estimation framework that we have presented uses an inherently batch algorithm designed to operate after data has been collected. An alternative approach, that is a natural extension of the framework, would allow for the algorithm to learn online, where techniques such as the recursive LS estimator \cite{Ljung99}, the incremental EM algorithm \cite{Neal98} or the stepwise EM \cite{Sato00,Cappe09} [see attached for details] may be used. These algorithms would facilitate dealing with non-stationary responses as well as providing a basis for optimal, model-based control, suggesting novel approaches to therapy.}

\section{Conclusion}
This paper has presented a framework for creating data-driven neural field models by developing a method for estimating intracortical connectivity and synaptic dynamics from electrophysiological data. In doing so, we have highlighted important aspects of the experiment design process, demonstrating the relationship between sensor bandwidth and the complexity of the field that can be inferred. We have derived a finite-order state-space representation of the continuum neural field equations that allows inference to be performed efficiently. Importantly, the mathematical formulation that we have presented scales well with the number of sensors. The major contribution of this framework is to formally bridge a gap between experimental and theoretical neuroscience in order to gain a better understanding of the mesoscopic dynamics of the brain. \dean{A validation of the estimation framework on real data is now required. This is the subject of planned future work.}
